On Optimality of Finite Element Integration

### 2015-12-11 TOMS_Paper/assembly.tex
Add stuff to proof plus minor fixes

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that characterize our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
We introduce the following notation and terminology
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
\item $c_o$: the cost (i.e., number of operations) of an expression $o$
\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
\end{itemize}

@@

We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$. The multilinear operands in a partition $P_i$ share a set of multilinear symbols, whereas there is no sharing across different partitions. We have to establish:
\begin{enumerate}
\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, then $P_i$ is scheduled before $P_j$, and
\item a scheduling strategy for each $P_i$
\end{enumerate}

We start with point 1). We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency. We then order $\mathbb{P}$ lexicographically, with the position of $P \in \mathbb{P}^j$ given by ${<}|\mathbb{P}^{j}|,\ |L_j| \sum_{o \in P} c_o{>}$ (and analogously for $P \in \mathbb{P}^k$). Intuitively, this simply means that operands appearing more often in the expression are handled first. We will later elaborate on this.

We now tackle point 2). For each partition, we have two options, as already explained in Section~\ref{sec:se-rln}. We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE; (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion. We adopt the following criteria:
\begin{enumerate}
\item Let $S_P$ be the set of multilinear symbols in $P$. If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops.
\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. Assume that $P$ depends on $[L_i, L_k]$. The increase in operation count at the level of the outer loops is: if (i) were applied, $|L_i| (|L_k| \sum_{o \in P} c_{o} + \sum_{o \in P} (n_{o} - 1))$; if (ii) were applied, $|L_i| \sum_{s \in P_S} (n_{s} - 1)$. We pick (i) or (ii) such that the increase is minimum. Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that they both minimize the cost of the innermost loop.
\end{enumerate}

@@

\subsubsection{Examples}
Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D. We observe that $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. Clearly, $\mathbb{P} = \lbrace P_0, P_1 \rbrace$, with $\mathbb{P}^k = \lbrace P_0 \rbrace$ and $\mathbb{P}^j = \lbrace P_1 \rbrace$. A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
\begin{figure}\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}

@@

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$. We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.

@@

// Find potential optimum such that memory constraints are honored
B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
for each b $\in$ B:
  for each (b$_S$, b$_P$) $\in$ b:
    if memory_required(b$_P$) $\geq$ $T_H$:
      continue
    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
(b$_S$, b$_P$) = min(op_count)

// Apply the transformations
pre_evaluate(P $\cup$ b$_P$)
eliminate_sharing(S $\cup$ b$_S$ $\cup$ P $\cup$ b$_P$)
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

@@

\begin{figure}\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}

@@

\begin{figure}[b]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}

@@

\subsection{Formalization}
\label{sec:proof}
The following proposition states that our approach guarantees quasi-optimality. The proof re-uses most of the concepts and explanations provided throughout the paper.

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form and the corresponding finite element integration loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$. Assume that pre-evaluating different monomials does not result in identical tables. Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
\end{Prop}
\begin{proof}
The function $\theta$ provides cost or gain of pre-evaluation. $\theta$ retrieves exact values, rather than estimates\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}. The estimate is actually for the whole form, rather than for its monomials.}, so $P$ and $Z$ are constructed correctly.

Let $\Lambda_{ops}$ denote the operation count (number of additions and products) in $\Lambda$. Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$, which holds if and only if
\begin{enumerate}
\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
\begin{enumerate}
\item \textit{pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$}
\item \textit{not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$}
\end{enumerate}
~
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two generic monomials $m_1, m_2 \in M$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. In a scenario in which $m_{T_{1}} \cap m_{T_{2}}$, we would have that at least one table would be assigned to the same temporary, with the transformation algorithm skipping a potential global optimum. By assumption, we neglect this possibility. The assumption is sustained by (i) the pathological nature of the scenario, (ii) the complexity inherent in handling it, and (iii) the fact that such a potential global optimum would probably be very close to the retrieved solution.
\item ...
\end{enumerate}

\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal.}\\
...
\end{enumerate}

>>>>

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural and algebraic properties that characterize our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
We introduce the following notation and terminology
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
\item $c_o^l$: the cost of an expression $o$ within a loop $L_l$; that is, the number of operations in $o$ times $|L_l|$.
\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
\end{itemize}

@@

We perform a second depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, with $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$. The multilinear operands within a partition share the same set of multilinear symbols, whereas there is no sharing across different partitions. We have to establish:
\begin{enumerate}
\item an ordering for $\mathbb{P}$, such that if $P^i \prec P^j$, then $P^i$ is handled before $P^j$, and
\item a scheduling strategy for each $P^i$
\end{enumerate}

We start with point 1). We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}_{j}$ and $\mathbb{P}_{k}$, which logically separates partitions based on loop dependency. We then build a reverse lexicographic ordering for $\mathbb{P}$, with the position of $P \in \mathbb{P}_l$, $l \in \lbrace j, k \rbrace$, given by ${<}|\mathbb{P}_{l}|,\ \sum_{o \in P} c_o^l{>}$. Intuitively, this simply means that operands appearing more often in the expression are handled first. We will later elaborate on this.

We now tackle point 2). For each partition, we have two options, as already explained in Section~\ref{sec:se-rln}. We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE; (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion. We adopt the following criteria:
\begin{enumerate}
\item Let $S_P$ be the set of multilinear symbols in $P$. If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops.
\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$. We pick (i) or (ii) such that the increase is minimum. Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that both will eventually minimize the innermost loop cost.
\end{enumerate}

@@

\subsubsection{Examples}
Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D. We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
\begin{figure}\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}

@@

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$. We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{algo:gamma}.

@@

// Find global quasi-optimum honoring memory constraints
B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
for each b $\in$ B:
  for each (b$_S$, b$_P$) $\in$ b:
    if memory_required(b$_P$) $\geq$ $T_H$:
      continue
    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
(b$_S$, b$_P$) = min(op_count)

// Apply the transformations
pre_evaluate(P $\cup$ b$_P$)
eliminate_sharing(S $\cup$ b$_S$ $\cup$ P $\cup$ b$_P$)
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{algo:gamma}
\end{figure}

@@

\begin{figure}[b]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}

@@

\begin{figure}[t]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}

@@

\subsection{Formalization}
\label{sec:proof}
The following proposition states that our approach guarantees quasi-optimality. The proof re-uses some of the concepts and explanations provided throughout the paper.

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form comprising a set of monomials $M$, and let $\Lambda$ be the corresponding finite element integration loop nest. Let $\Gamma$ be the transformation algorithm. Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$. Assume that pre-evaluating different monomials does not result in identical tables. Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
\end{Prop}
\begin{proof}
We first observe that the cost function $\theta$ predicts the exact gain/loss in monomial pre-evaluation, so $P$ is always correct by construction.

Let $c_\Lambda$ denote the operation count in $\Lambda$. Let $\Lambda'_I \subset \Lambda'$ be the subset of all innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ such that $c_{\Lambda''_I} < c_{\Lambda'_I}$. This holds if and only if
\begin{enumerate}
\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
\begin{enumerate}
\item \textit{pre-evaluating any $Z_P \subseteq Z$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\item \textit{not pre-evaluating any $P_Z \subseteq P$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\end{enumerate}
\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces (at least) quasi-optimal loop nests.}\\
\end{enumerate}

We discuss these points separately

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two generic monomials $m_1, m_2 \in Z$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $p \in P$ and $z \in Z$ two monomials that share some subsets of multilinear symbols. If, as by assumption $\theta(p)$ suggests, $p$ will be pre-evaluated, there may be a potential gain in sharing elimination going lost, which potentially leads to a locally optimal solution -- instead of a global one. We avoid this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item ...
\end{enumerate}
